---
title: "Dubious ways to take the exponential of a matrix"
author: "Adam Sardar"
date: "26/05/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---


 There a number of means of taking the exponential of a matrix. This is a script exploring
 
 
 
```{r}

devtools::load_all()
library(Matrix)
```




```{r}

fbMod <- fitBetaUniformMixtureDistribution(siTAZdiffex_HFL1_diffexDT$siTAZ1_pValue)

testPvalues <- rbetauniform(3, a = fbMod@a, lambda = fbMod@lambda)

nValues <- length(testPvalues)
fittedBetaShape <- fbMod@a
uniformProportion <- fbMod@lambda
            
testStatistic <- sum(-log(testPvalues))
            
transitionMatrix <- constructHyperexponentialSumTransitionMatrix(nValues,
                                                                 uniformProportion,
                                                                 fittedBetaShape)
            

```

# QR decomp

```{r}

qrDecomp <- qr(as.matrix(transitionMatrix))

QRmat <- solve(transitionMatrix, diag(diag(transitionMatrix))) #Vmat


QRmat %*% expm(Dmat) %*% solve(Rmat, Qmat)


expm(transitionMatrix)

Rmat <- qr.R(qrDecomp)
Qmat <- qr.Q(qrDecomp)

Tmat <- Qmat %*% transitionMatrix %*% t(Qmat)

Dmat <- diag(diag(transitionMatrix))



Qmat %*% Rmat %*% Dmat

transitionMatrix %*% Qmat %*% Rmat

EIGS <- eigen(Tmat)

Vmat <- EIGS$vectors

solve(qrDecomp)


```

# Schur

```{r}

schurDecomp <- Schur(transitionMatrix, vectors = TRUE)

```


Supposedly computation of the exponential of an upper-triangular matrix is easy - the Shur-Partlett method. Given the sparsity and regularly repeated elements, we should be able to compute it very quickly.

Doubly so when we consider that we actually only care about a handful of elements (since we start at two positions, so we only need the top two rows of exp(A) ). Very easy to do using recursion

## Truncated series

https://books.google.co.uk/books?id=2Wz_zVUEwPkC&pg=PA86&lpg=PA86&dq=parlett+triangular&source=bl&ots=pUB4drJJU1&sig=ACfU3U3UF-OSnrkQkjU8U7fPbpGVU91gGA&hl=en&sa=X&ved=2ahUKEwi49t2KjtfpAhWHa8AKHV9kC5QQ6AEwAnoECAcQAQ#v=onepage&q=parlett%20triangular&f=false

```{r}

T1 <- constructHyperexponentialSumTransitionMatrix(1, uniformProportion, fittedBetaShape)
T2 <- constructHyperexponentialSumTransitionMatrix(2, uniformProportion, fittedBetaShape)
T3 <- constructHyperexponentialSumTransitionMatrix(3, uniformProportion, fittedBetaShape)
T4 <- constructHyperexponentialSumTransitionMatrix(4, uniformProportion, fittedBetaShape)


D1 <- diag(diag(T1))
D2 <- diag(diag(T2))
D3 <- diag(diag(T3))
D4 <- diag(diag(T4))

M1 <- T1; diag(M1) <- 0
M2 <- T2; diag(M2) <- 0
M3 <- T3; diag(M3) <- 0
M4 <- T4; diag(M4) <- 0

```


```{r}

library(microbenchmark)

library(expoRkit)



largeTransition <- constructHyperexponentialSumTransitionMatrix(5, uniformProportion, fittedBetaShape)

expoRkit::expv(largeTransition, v = rep(1,2*5),
                                    t = 1)

transitionMatrixTriag <-  as(largeTransition, "triangularMatrix")

transitionMatrixTriagDropped0 <- drop0(largeTransition)

transitionMatrixTriagDropped0Triag <- as(transitionMatrixTriagDropped0, "triangularMatrix")
X <- transitionMatrixTriagDropped0Triag


X.csc <- new("matrix.csc", ra = X@x,
                           ja = X@i + 1L,
                           ia = X@p + 1L,
                           dimension = X@Dim)


ness <- SparseM_coo_to_REXPOKIT_coo (X.csc)


SparseM_coo_to_REXPOKIT_coo(ness)

microbenchmark(
  Matrix::expm(transitionMatrixTriag),
  Matrix::expm(largeTransition),
  Matrix::expm(transitionMatrixTriagDropped0), 
  Matrix::expm(transitionMatrixTriagDropped0Triag))


```

```{r}

nValues <- 50

EXPOKITmat <- data.table(ia = c(seq(1,2*nValues), seq(1,2*nValues-1), seq(1,2*nValues-2), seq(1,2*nValues-3)) , 
                 ja = c(seq(1,2*nValues), seq(2,2*nValues), seq(3,2*nValues), seq(4,2*nValues)) ,
                 a = c(rep(c(-1,-fittedBetaShape), times = nValues), #diagonal
                         c(rep(c(0,fittedBetaShape*uniformProportion), times = (nValues-1)),0), # diagonal+1
                         rep(c(uniformProportion,fittedBetaShape*(1-uniformProportion)), times = (nValues-1)), # diagonal+2
                         c(rep(c((1-uniformProportion),0), times = (nValues-2)),(1-uniformProportion))  ) ) # diagonal+3 


testTranisitionMat <- constructHyperexponentialSumTransitionMatrix(nValues = nValues, uniformProportion, fittedBetaShape)


EXPOKITmat <- EXPOKITmat[abs(a) > 1E-10]

n = nValues
m = n-1
ideg = 6
lwsp = as.integer(n * (m + 2) + 5 * (m + 2) ^ 2 + ideg + 1)
liwsp = max(m + 2, 7)
iwsp = integer(length = liwsp)



testTranisitionMatTriag <-  as(testTranisitionMat, "triangularMatrix")



expoRkit::expv(testTranisitionMat, v = rep(1,2*nValues), t = c(1,2,3))

microbenchmark::microbenchmark(
  
  Matrix::expm(testTranisitionMat),
  
    Matrix::expm(testTranisitionMatTriag),

  expm::expm.Higham08(testTranisitionMat, balancing = FALSE),
  
  expoRkit::expv(testTranisitionMat, v = rep(1,2*nValues), t = 1 ),
  
  rexpokit:::expokit_mydgexpv_wrapper(n = n, 
                                    m = m,
                                    
                                    v = rep(1,2*nValues),
                                    t = 1, 
                                    
                                    tol = 1E-6,
                                    anorm = 20,
                                    
                                    wsp = double(length = lwsp),
                                    iwsp = iwsp,
                                    lwsp = lwsp,
                                    liwsp = liwsp,
                                    
                                    ia = EXPOKITmat$ia,
                                    ja = EXPOKITmat$ja,
                                    a = EXPOKITmat$a,
                                    nz = nrow(EXPOKITmat) )
)




microbenchmark::microbenchmark(
 Matrix::expm(testTranisitionMat),
 Matrix::expm(drop0(testTranisitionMat)) %*% rep(1, 2*nValues),
 expoRkit::expv(testTranisitionMat, v = rep(1,2*nValues), t = 1 ),
 expoRkit::expv(drop0(testTranisitionMat), v = rep(1,2*nValues), t = 1 )
)




            initialProb <- Matrix(0,ncol = ncol(testTranisitionMat), nrow = 1)
            initialProb[1:2] <- c(uniformProportion, 1-uniformProportion)
            
            sum(initialProb %*% Matrix::expm(5*testTranisitionMat))


tmpmat_in_REXPOKIT_coo_fmt 

rexpokit::expokit_dgexpv_Qmat()


rexpokit:::expokit_mydgexpv_wrapper(n = n, 
                                    m = m,
                                    
                                    v = rep(1,2*nValues),
                                    t = c(1,2,3,4), 
                                    
                                    tol = 1E-6,
                                    anorm = 20,
                                    
                                    wsp = double(length = lwsp),
                                    iwsp = iwsp,
                                    lwsp = lwsp,
                                    liwsp = liwsp,
                                    
                                    ia = EXPOKITmat$ia,
                                    ja = EXPOKITmat$ja,
                                    a = EXPOKITmat$a,
                                    nz = nrow(EXPOKITmat) )


```


Looks like expoRkit is the right choice. Correct use of S4 
