---
title: "Processing GSE50588 to produce validation sets"
author: "Adam Sardar"
date: "19 November 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

# TODO detail motivation further

[Cusanovich et al.](http://dx.plos.org/10.1371/journal.pgen.1004226) have prepared a fantastic dataset of transcription factors knocked down with siRNA in a lymphoblastoid cell line. This is a perfect dataset for producing a gold standard form the metaDEGth algorithm.

"201 samples were analyzed - 57 transcription factors were knocked down in triplicate in the same cell line, 2 factors were knocked down in 6 replicates each and an additional 18 control knockdowns (siRNA not targeting any known gene) were analyzed"

The authors use Illumina HumanHT-12 V4.0 expression beadchips, so a pipeline that makes use of the detection P-value should be employed. Luckily limma solves the problem for us!

# Import data

Download the unnormalised zipped expression data directly from GEO

```{r Read in data directly from GEO}

library(data.table)

GSE50588_unnormalised_RAW_DT <- fread(
  cmd = "curl ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE50nnn/GSE50588/suppl/GSE50588_non-normalized.txt.gz | zcat",
  check.names = TRUE)
```

Build an annotation frame for the experimental system

```{r Sample metadata extraction and clean up}

library(stringr)

samples_GSE50588_metadataDT <- data.table(
  sampleNames = names(GSE50588_unnormalised_RAW_DT) )

samples_GSE50588_metadataDT[, TFknockdown := shift(sampleNames)]
samples_GSE50588_metadataDT[sampleNames %like% "Detect", sampleNames := str_c(TFknockdown,"_DetectionPval")]
samples_GSE50588_metadataDT[!sampleNames %like% "detect", TFknockdown :=  sampleNames]
samples_GSE50588_metadataDT[, TFknockdown := str_remove(TFknockdown, "_\\d(_rep)?")]

samples_GSE50588_metadataDT[sampleNames %like% "NS", TFknockdown := "CONTROL"]

samples_GSE50588_metadataDT[,
                            type := ifelse(str_detect(sampleNames,"Pval"),"pValue","expression"),
                            by = sampleNames]
samples_GSE50588_metadataDT[sampleNames == "ID_REF", type := "ID"]

samples_GSE50588_metadataDT[type == "expression",.N ,by = TFknockdown][order(N)]
```

# Background correct and quantile normalise

We shall make use of the  `neqc` function in the `limma` package, which is designed for use with Illumina BeadChips; it performs background correction followed by quantile normalisation across all samples, using detection P-values provided by the BeadArray detector.

```{r Normalise Raw Measurements}

library(limma)

setnames(GSE50588_unnormalised_RAW_DT, samples_GSE50588_metadataDT$sampleNames)

exprsMat <- GSE50588_unnormalised_RAW_DT[, 
                                    as.matrix(.SD, rownames = ID_REF),
                    .SDcols = samples_GSE50588_metadataDT[type == "expression", sampleNames] ]

detectionPvalueMat <- GSE50588_unnormalised_RAW_DT[,
                                    as.matrix(.SD, rownames = ID_REF),
                    .SDcols = samples_GSE50588_metadataDT[type == "pValue", sampleNames] ]

GSE50588_expObj <- neqc(x = exprsMat,
                        detection.p = detectionPvalueMat)
```

# Map Ilumina probe IDs

GEO provides an annotation file detailing [GPL10558](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL10558), corresponding to the HumanHT-12 v4 Expression BeadChip used in the experiment.

```{r Construct annotation frame}

GPL10558_annot_DT <- fread(cmd = "curl ftp://ftp.ncbi.nlm.nih.gov/geo/platforms/GPL10nnn/GPL10558/suppl/GPL10558_HumanHT-12_V4_0_R2_15002873_B.txt.gz | zcat", skip = 8, header = TRUE, fill = TRUE)

GPL10558_annot_DT <- unique(GPL10558_annot_DT[Probe_Id != "",.(Probe_Id, RefSeq_ID, Entrez_Gene_ID, Symbol ,Definition)])

setkey(GPL10558_annot_DT, "Probe_Id")
```

# Fit limma model and estimate differential expression

A linear model fit first requires a design matrix detailing what contrasts can be drawn between which experiments.

```{r Produce contrast and design matricies}

assayedTFs <- samples_GSE50588_metadataDT[type == "expression" & TFknockdown != "CONTROL", unique(TFknockdown)]
          
GSE50588_designMat <- model.matrix(sampleNames ~ 0 + TFknockdown,
                                   samples_GSE50588_metadataDT[type == "expression",.(sampleNames,TFknockdown)])

colnames(GSE50588_designMat) %<>% str_remove("TFknockdown")
row.names(GSE50588_designMat) <- samples_GSE50588_metadataDT[type == "expression", sampleNames]


GSE50588_TFcontrastMat <- makeContrasts(
              contrasts = str_c("CONTROL-",assayedTFs),
              levels = GSE50588_designMat)
```

We are now ready to build models and compare contrasts (siRNA knockdowns). The plan is to 

```{r Fit model and estimate differential expression}

library(purrr)

GSE50588_lmFit <- lmFit(GSE50588_expObj,
                        design = GSE50588_designMat) %>%
                  contrasts.fit(contrasts = GSE50588_TFcontrastMat) %>%
                  eBayes

TF_siRNA_knockdowns_diffex <- map(
  assayedTFs,
  ~{ diffexDT <- data.table(topTable(GSE50588_lmFit, coef = which(.x==assayedTFs), number = Inf),
                            keep.rownames = TRUE )
     
    setnames(diffexDT,
             c("rn","logFC","P.Value"),
             c("HumanHT12v4probeID", str_c(.x,c("_logFC","_pValue")) ))
    
    setkey(diffexDT, "HumanHT12v4probeID")
    
    return(diffexDT[,.SD,.SDcols = c("HumanHT12v4probeID", str_c(.x,"_logFC") , str_c(.x,"_pValue"))])
  })

TF_siRNA_knockdowns_GSE50588_summary <- Reduce(function(x,y){x[y]}, TF_siRNA_knockdowns_diffex) %>%    
                                .[GPL10558_annot_DT, , on = .(HumanHT12v4probeID == Probe_Id)]

save(TF_siRNA_knockdowns_GSE50588_summary,
     file = "./data/TF_siRNA_knockdowns_GSE50588_summary.RData",
     compress = "xz")

object.size(TF_siRNA_knockdowns_GSE50588_summary) %>% print(units = "Mb")

```

# TODO 40Mb data files are far too big. Cull somehow.

# TODO - a few volcano plots with some regulons highlighted
