---
title: "Aquired resistence to third line TKI in NCI-H1975 cell lines"
author: "Adam Sardar"
date: "9 December 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

[GSE37699](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE37699)

A pretty straighforward experiment; the authors prepare cell lines that are resistent to WZ4002 (a third generation TKI) and then compare the expression of the two cell lines. Prepared in triplicate and measured on Affymetrix Human Genome U133A 2.0 Arrays.

From the study blurb:

"Our study identifies ERK signaling as a mediator of resistance to irreversible pyrimidine EGFR inhibitors in EGFR T790M-bearing cancers. We further provide a therapeutic strategy to both treat and prevent the emergence of this resistance mechanism."

Should be a good reporter assay for testing different datasets.

```{r}

library(data.table)
library(GEOquery)
library(purrr)
library(stringr)
library(magrittr)

parsedGSE37699SOFT <- getGEO(GEO = "GSE37699", GSEMatrix = FALSE, AnnotGPL = FALSE, getGPL = TRUE)

GPL571_table <- data.table(parsedGSE37699SOFT@gpls$GPL571@dataTable@table)

GSE37699_metadata <- map_dfr(parsedGSE37699SOFT@gsms,
    ~ as.data.table(.x@header) )

GSE37699_metadata[, treatment := str_remove(title, "_rep\\d")]
GSE37699_metadata[, filename := str_remove(supplementary_file, "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM925nnn/GSM\\d+/suppl/")]

GSE37699_metadata[, study_name := str_remove(title, "NCI-")]

```


```{r}
library(affy)

GSE37699_CELfiles <- list.files("~/GSE37699/CELfiles/", pattern = "CEL", full.names = TRUE)

setwd("~/GSE37699/CELfiles/")
GSE37699_affy <- ReadAffy()

GSE37699_normalised <- rma(GSE37699_affy)

GSE37699_exprs <- exprs(GSE37699_normalised) # I prefer to just work with matricies. It's really explicit and simple

setkey(GSE37699_metadata, filename)

colnames(GSE37699_exprs) <- GSE37699_metadata[colnames(GSE37699_exprs), study_name]

```

Build a design and contrast matrix

```{r Produce contrast and design matrices}

GSE37699_designMat <- model.matrix(study_name ~ 0 + treatment,
                                   GSE37699_metadata[filename %like% "CEL"])

colnames(GSE37699_designMat) %<>% str_remove("treatment") %>% str_remove("NCI-")
row.names(GSE37699_designMat) <- GSE37699_metadata[filename %like% "CEL", study_name]

GSE37699_contrastMat <- makeContrasts(
              contrasts = c("H1975_WZR6-H1975_parental"), #challenge - control
              levels = GSE37699_designMat)

```


```{r RMA}
library(limma)


GSE37699_lmFit <- lmFit(GSE37699_exprs,
                        design = GSE37699_designMat) %>%
                  contrasts.fit(contrasts = GSE37699_contrastMat) %>%
                  eBayes


WZ4002_adrH1975_diffexDT <- data.table(topTable(GSE37699_lmFit, number = Inf, adjust.method = "fdr",
                                                coef = "H1975_WZR6-H1975_parental"), keep.rownames = TRUE )

WZ4002_adrH1975_diffexDT %<>% 
  merge(GPL571_table[,.(ID, geneSymbol = `Gene Symbol`, geneID = ENTREZ_GENE_ID)], by.x = "rn", by.y = "ID")


setnames(WZ4002_adrH1975_diffexDT,
         c("rn","logFC","P.Value"),
         c("HG_U133A_2_ID", "H1975_logFC", "H1975_pValue"))


WZ4002_adrH1975_diffexDT <- WZ4002_adrH1975_diffexDT[,.SD, .SDcols = !c("AveExpr","t","B","adj.P.Val")]

setcolorder(WZ4002_adrH1975_diffexDT, c("HG_U133A_2_ID","geneSymbol","geneID"))

```



```{r}

WZ4002_adrH1975_diffexDT %>%
  ggplot(aes(x = H1975_logFC, y = -log10(H1975_pValue))) +
  geom_point()

setorder(WZ4002_adrH1975_diffexDT, H1975_pValue)

save(WZ4002_adrH1975_diffexDT, file = "./data/WZ4002_adrH1975_diffexDT.RData", compress = "xz")

```



