---
title: "WZ4002 ADR enrchment"
author: "Adam Sardar"
date: "04/06/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---




Study using BioPlanet

# Should see MAPK1 and/or ERK. EGFR too.

```{r}

WZ4002_adrH1975_diffexDT[,qplot(H1975_pValue)]

WZ4002DT <- WZ4002_adrH1975_diffexDT[,.SD[H1975_pValue == min(H1975_pValue)],by = geneID]

tkiBUM <- fitBetaUniformMixtureDistribution(WZ4002DT$H1975_pValue)

print.BetaUniformPlots(plot(tkiBUM))


tkiWZ4002enrichDT <- metaDEG(WZ4002DT,
        bioplanetPathwaysDT,
        tkiBUM,
        pValAttr = "H1975_pValue",
        geneSetNameAttr = "BioPlanetName",
        geneSetMembersAttr = "geneID" )

WZ4002_adrH1975_diffexDT

```

And we do! ERK is more of a broad scale collection of pathways; any growth factor will activate it. Here we have a plausible mechanism that would unlock it.

https://en.wikipedia.org/wiki/Extracellular_signal-regulated_kinases

The most that anyone can do is propose a future route for experiments and analysis. Type II error is extremely expensive and false positives are far too often toleratred to find a promising result which vanishes.

This could be the case with metaDEGth - but so it is provably so with wilcoxen, GSEA and fisher.


```{r}

erlotinibDT <- erlotinib_adrHCC827_diffexDT[,.SD[ER3_pValue == min(ER3_pValue)],by = geneID]

erloE3BUM <- fitBetaUniformMixtureDistribution(erlotinibDT$ER3_pValue)

print.BetaUniformPlots(plot(erloE3BUM))

erlotinibEnrichDT <- metaDEG(erlotinibDT,
        bioplanetPathwaysDT,
        erloE3BUM,
        pValAttr = "ER3_pValue",
        geneSetNameAttr = "BioPlanetName",
        geneSetMembersAttr = "geneID" )

```


```{r}

library(fgsea)

hallmark_pathways <- gmtPathways("/mnt/c/broad_genesets/h.all.v7.0.entrez.gmt")
hallmark_pathways_DT <- map2_dfr(hallmark_pathways, names(hallmark_pathways), ~ data.table(name = .y, geneID = .x))


cannonical_pathways <- gmtPathways("/mnt/c/broad_genesets/c2.cp.v7.0.entrez.gmt")
cannonical_pathways_DT <- map2_dfr(cannonical_pathways, names(cannonical_pathways), ~ data.table(name = .y, geneID = .x))

chemical_genetic_peturbations <- gmtPathways("/mnt/c/broad_genesets/c2.cgp.v7.0.entrez.gmt")
CGP_geneSets_DT <- map2_dfr(chemical_genetic_peturbations, names(chemical_genetic_peturbations), ~ data.table(name = .y, geneID = .x))



WZ4002_geneIDexpress <- WZ4002_adrH1975_diffexDT[!is.na(geneID), .SD[H1975_pValue == min(H1975_pValue, na.rm = T)][1] , by = geneID]


WZ4002_betaUniformModel <- fitBetaUniformMixtureDistribution(WZ4002_geneIDexpress$H1975_pValue, nStarts = 20)

WZ4002_geneIDexpress[, betaUnifScore_FDR0.05 := betaUniformScore(H1975_pValue, WZ4002_betaUniformModel, FDR = 0.05)]

noiseFractionUpperBound(WZ4002_betaUniformModel)

####


cgpSetsDT <- WZ4002_geneIDexpress[ unique(cannonical_pathways_DT[, .(pathway = name, geneID)]), , on = "geneID"][!is.na(H1975_pValue),.(pValueSet = list(H1975_pValue), geneSet = list(geneSymbol), scoreSum = sum(betaUnifScore_FDR0.05)), by = pathway]

cgpSetsDT[, betaUniformMixtureP := betaUniformPvalueSumTest(pValueSet[[1]], WZ4002_betaUniformModel), by = pathway]
cgpSetsDT[, fishersP := fishersPvalueSumTest(pValueSet[[1]]), by = pathway]


cgpSetsDT[pathway %like% "BIOCARTA"][p.adjust(betaUniformMixtureP, "fdr") < 0.01] 
cgpSetsDT[p.adjust(fishersP, "fdr") < 0.01]

cgpSetsDT[scoreSum > 0]



```
